Management induced changes of soil organic carbon on global croplands

### 2020-12-14 SOCbudgetPaper/SOCbudget.tex
working pdf complilation

\begin{abstract}
Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. It is three times larger than the vegetation carbon pool and more than twice as big as the atmospheric carbon pool, when looking into the first meter of the Earth's soil profile. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with the IPCC Tier 2 steady-state soil model to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have around 26 GtC relative to a counterfactual natural state in 1975. Yet, within the period 1975--2010 this SOC debt has been decreasing again by a net quantity of 4 Gt SOC, which can be mainly traced back to an increased input of C in crop residues due to higher crop productivity. We also find that SOC is very sensitive to management decisions such as residue returning indicating the necessity to incorporate better management data in soil model simulations.
\end{abstract}

@@

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions (\citep{friedlingstein_global_2019}). While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest (\citep{strassmann_simulating_2008}, \citep{betts_climate_2015}). BKMs are designed to estimate LUC realted emissons but ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization, and agricultural management (\citep{friedlingstein_global_2019}, \citep{houghton_carbon_2012}, \citep{hansis_relevance_2015}).

@@

The objective of our study is to provide the first global, spatially explicit SOC and SOC debt map that considers spatially explicit and time-variant historical agricultural management. To achieve this objective we create and provide a comprehensive dataset of the global gridded management time series data, including crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We simulate SOC stocks, dynamics and the SOC debt for 1975--2010. Using a scenario analysis, we decompose the contribution of different management activities. Our analysis highlights the importance of considering agricultural management for improving estimates of SOC stocks. We conclude, that historic agricultural management let to increasing SOC stocks in the most recent decades.
\newpage

@@

In Sect. \ref{sec:carbonbudget} we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}). We additionally describe how we configured and extended the steady-state method (for model code see \citep{karstens_mrsoil_2020}). In Sect. \ref{sec:tier1} we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines (\citep{ipcc_eggleston_2006}, \citep{calvo_buendia_ipcc_2019}).
Section \ref{sec:agrimanagement} provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see \citet{mrcommons_2020}). In Sect. \ref{sec:scenarios} we define the scenarios used to complement our historic model results.

@@

Following the Tier 2 steady-state approach of the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}; referred to as \textit{steady-state method} in the following), we estimate soil organic carbon (\(SOC\)) stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current \(SOC\) state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period:
(1) We calculate annual land-use type-specific steady states and decay rates for \(SOC\) stocks (Sect. \ref{sec:steadystates});
(2) we account for land conversion by transferring \(SOC\) from and natural vegetation (Sect. \ref{sec:carbontransfer}),
(3) we estimate \(SOC\) stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \ref{sec:totalsoc}).
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \ref{sec:initsoc}).

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the steady-state method for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model \citep[(][]{parton_analysis_1987}). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below and above ground residues left or returned to the field (see Sect. \ref{sec:residues}) and manure inputs (see Sect. \ref{sec:livstmanure}); for natural vegetation litterfall (including fine root turnover) (\citep{schaphoff_lpjml4_2018}) is the only source of carbon inflow to the soil. Following the IPCC guidelines \citep{calvo_buendia_ipcc_2019}, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective \(SOC\) sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \ref{tab:datasourceinputs}.

  \begin{table*}[h]
  \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
  \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.40\textwidth}}
  \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
  \middlehline
  \multirow{3}{*}{Cropland} & above ground residues & \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                      \cite{faostat_faostat_2016}, \\
                                                                      \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                      \cite{weindl_livestock_2017} \end{minipage}} & 
                                                      \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                      lg/c generic values based on Table 5.5B, 5.5C \\
                                                                      in \cite{calvo_buendia_ipcc_2019},\\
                                                                      crop-type specific n/c-ratios from
                                                                      \cite{bodirsky_mrcommons_2020}\end{minipage}} \\
                            & below ground residues &  &  \\ 
                            & manure                &  &  \\ 
                            \hline
  Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                            & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

@@

Note that the absolute changes \(\Delta SOC\) can be also interpreted as the SOC debt (\citep{sanderman_soil_2017}) due to human cropping activities; whereas relative changes \(F^{\mathrm{SCF}}\) can be considered stock change factors as defined within the IPCC guidelines of 2006 \citep{eggleston_ipcc_2006}. Moreover, \(\Delta SOC\) is equivalent to the negated cumulative SOC component of human land use change emission (\citep{pugh_simulated_2015}).

@@

We compile country-specific FAO production and cropland statistics \citep{faostat_faostat_2016} to a harmonized and consistent data set. The data is prepared in 5-year time steps from 1965 to 2010, which, together with the spin-up phase, restricts our analysis to the time span from 1975 to 2010. For all the following data, if not declared differently, we interpolate values linearly between the time steps and keep them constant before 1965.

@@

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL \citep{schaphoff_lpjml4_2018} as well as half-degree cropland patterns (see Sect. \ref{sec:landuse}). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production \citep{faostat_faostat_2016}. Note that by using physical cropland areas we account for multiple crop harvest events as well as for fallow land.

@@

Following the IPCC guidelines, we split the harvest index into a production (\(HI^{\mathrm{prod}}\)) and an area dependent (\(HI^{\mathrm{area}}\)) fraction \citep{eggleston_ipcc_2006}. Note that deviating from \citep{bodirsky_n2o_2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all \(BGR\) are left in the soil, whereas \(AGR\) can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

A fixed share of the \(AGR\) is assumed to be burned on field depending on the per-capita income of the country. Following \citep{smil_nitrogen_1999}) we assume a burn share of \(25\%\) for low-income countries according to World Bank definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80--90\% of the carbon in the crop residues burned in the fields is lost within the combustion process \citep{eggleston_ipcc_2006}.

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in \citep{bodirsky_mrcommons_2020}). Residue feed demand for five different livestock groups is based on country-specific feed baskets (see \citep{weindl_livestock_2017}), that differentiate between the residue groups and take available \(AGR\) biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5\% and a fuel-share of 10\% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining \(AGR\) as well as all \(BGR\) are expected to be left on the field. We limit high returning rates at \(10\unit{tC\,ha}^{-1}\) to correct for outliers.

@@

To disaggregate country level FAOSTAT livestock production data to half-degree resolution, we use the following rule-based assumptions, drawing from the approach of \citep{robinson_mapping_2014} and applying feed basket assumptions based on a revised methodology from \citep{weindl_livestock_2017}. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
Due to great feed demand of ruminants, we assume that ruminant livestock is located where the production of feed occurs to minimize transport of feed. We distinguish between grazed pasture, which is converted into livestock products in extensive systems, and primary-crop feed stuff, which we consider to be consumed in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to distinguish between intensive and extensive production systems. For low-income countries, we assume only extensive production systems. We locate them according to the share of built-up areas based on the asumption that these animals are held in subsistence or small-holder farming systems with a high labor-per-animal ratio. Intensive production associated with high-income countries, is distributed within a country using the share of primary-crop production, assuming that feed availability is the most determining factor for livestock location. For middle-income countries we split the livestock production into extensive and intensive systems based on the per-capita income.

Manure production and management is based on a revised methodology of \citep{bodirsky_n2o_2012} and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets \citep{weindl_livestock_2017}, assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed c/n ratios (given by \citep{calvo_buendia_ipcc_2019}).
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25\% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing.
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates depending on the different systems and the associated nitrogen loss rates as specified in \citep{bodirsky_n2o_2012} (see Sect. \ref{append:TableclossAWMS} for more details). We limit high application shares at \(10\unit{tC\,ha}^{-1}\) to correct for outliers.

@@

The LUH2v2 \citep{hurtt_harmonization_2020} data set provides irrigated fractions for their cropland subcategories. We sum up irrigation area shares for all crop groups within a grid cell, and calculate the water effect coefficient \(wat\) on decay rates using these shares to compute the weighted mean between rainfed and irrigated \(wat\) factors. As a result \(wat\) is the same for all crop groups within a grid cell. Furthermore, we suppose the irrigation effect to be present for all 12 months of a year in a (partly) irrigated grid cells, since we do not have crop group specific growing periods at hand. In addition, we expect, either water limitations to be no problem during the off-season (causing our assumption to be a good approximation), or irrigated growing periods to span over the whole year in water-restricted areas.

@@

To single out the impact of tillage practices, residue and manure inputs, we defined scenario with constant values for the three drivers: (1) In the \textit{constTillage} scenario the adoption of no-tillage practices are neglected (adoption starts in 1974 according to the available data set); (2) the \textit{constResidues} and the \textit{constManure} scenario assume constant input rates from residues resp. manure (in \(\tfrac{t}{ha}\)) at the level of 1975 onwards. Within the \textit{constResidue} scenario at least three effects overlay each other: (1) yields and with them residue biomass increase due to productivity gains, (2) rates of residue left or returned to fields are raising, and (3) shifts of crop groups change the amount of residue biomass due to crop-group specific harvest index values. The \textit{constManagement} combines all three scenarios.

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short lg/c) and nitrogen-to-carbon (short n/c) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines (\citep{calvo_buendia_ipcc_2019}), but for woody biomass we used the CENTURY configuration file (\citep{century_model_2000}) for trees as parameterization were not given. Whereas lg/c ratios ranges between \(0.09-0.35\) with an average of \(0.24\) (default value) for different tree compartments and different tree types, n/c ratios range between \(0.001-0.05\) with an average of \(0.015\) (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table \ref{tab:scenlitterpnv}.

 \begin{table*}[h]
 \caption{Scenarios for sensitivity analysis on litter parameterization}
 \begin{tabular}{l l l l}
 \tophline
  Scenario name & plant type & n/c & lg/c \\
  \middlehline
  \textit{LitterPNV-Mean} & mix over all tree types & 0.015  & 0.25 \\
  \textit{LitterPNV-Grass} & perennial grasses     & 0.029  & 0.11 \\
  \textit{LitterPNV-AASS}  & Arctic alpine subshrub           & 0.022  & 0.22 \\
  \textit{LitterPNV-BLEAF} & Temperate broadleaf evergreen    & 0.013  & 0.27 \\
  \textit{LitterPNV-BRCF}  & Boreal coniferous forest         & 0.013  & 0.17 \\
  \textit{LitterPNV-MTCF}  & Maritime coniferous forest       & 0.0094 & 0.22 \\
  \textit{LitterPNV-TREF}  & Tropical evergreen forest        & 0.016  & 0.30 \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:scenlitterpnv}
 \end{table*}

@@

\hypertarget{agricultural-management-matters}{%
\subsection{Agricultural management matters}\label{agricultural-management-matters}}

This study shows that spatially explicit and time-variant historical agricultural management data changes estimations of states and trends of global SOC and SOC debt considerably compared to constant management assumptions. This result remains robust to variations of central model parameters and initialization stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to net reduction of the SOC debt.

The evaluation of SOC stock gains and losses, however, is complex and has several dimensions as climatic and anthropogenic effects overlap. If defining the SOC debt of 1975 the baseline and measuring land-use emissions on croplands as the difference between a potential natural state and the state under human interventions (see \citep{pugh_simulated_2015}), global croplands have acted as a carbon sink according to our study. Annual C sequesteration rates of 0.2 per mille are well below the promoted 4 per mille (\citep{minasny_soil_2017}), indicating that productivity gains on historic levels alone are not enough to meet ambitious climate targets.

>>>>

\begin{abstract}
Soil organic carbon (SOC) is one of the largest terrestrial carbon stocks on Earth. It is three times larger than the vegetation carbon pool and more than twice as big as the atmospheric carbon pool, when looking into the first meter of the Earth's soil profile. SOC has been depleted by anthropogenic land cover change and agricultural management. However, the latter has so far not been well represented in global carbon stock assessments. While SOC models often simulate detailed biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated at the global scale. Here we develop a spatial explicit data set for agricultural management on cropland, considering crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We combine it with the IPCC Tier 2 steady-state soil model to create a half-degree resolution data set of SOC stocks and SOC stock changes for the first 30 cm of mineral soils. We estimate that due to arable farming, soils have lost around 26 GtC relative to a counterfactual natural state in 1975. Yet, within the period 1975--2010 this SOC debt has been decreasing again by a net quantity of 4 Gt SOC, which can be mainly traced back to an increased input of C in crop residues due to higher crop productivity. We also find that SOC is very sensitive to management decisions such as residue returning indicating the necessity to incorporate better management data in soil model simulations.
\end{abstract}

@@

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions (\citep{friedlingstein_global_2019}). While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest (\citep{strassmann_simulating_2008}, \citep{betts_climate_2015}). BKMs are designed to estimate LUC related emissons but ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization, and agricultural management (\citep{friedlingstein_global_2019}, \citep{houghton_carbon_2012}, \citep{hansis_relevance_2015}).

@@

The objective of our study is to provide the first global, spatially explicit SOC and SOC debt map that considers spatially explicit and time-variant historical agricultural management. To achieve this objective we create and provide a comprehensive data set of the global gridded management time series data, including crop production levels, residue returning rates, manure application, and the adoption of irrigation and tillage practices. We simulate SOC stocks, dynamics and the SOC debt for 1975--2010. Using a scenario analysis, we decompose the contribution of different management activities.
\newpage

@@

In Sect. \ref{sec:carbonbudget} we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}). We additionally describe how we configured and extended the steady-state method (for model code see \citep{karstens_mrsoil_2020}). In Sect. \ref{sec:tier1} we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines (\citep{eggleston_ipcc_2006}, \citep{calvo_buendia_ipcc_2019}).
Section \ref{sec:agrimanagement} provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see \citep{bodirsky_mrcommons_2020}). In Sect. \ref{sec:scenarios} we define the scenarios used to complement our historic model results.

@@

Following the Tier 2 steady-state approach of the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}; referred to as \textit{steady-state method} in the following), we estimate soil organic carbon (\(SOC\)) stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current \(SOC\) state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period:
(1) We calculate annual land-use type-specific steady states and decay rates for \(SOC\) stocks (Sect. \ref{sec:steadystates});
(2) we account for land conversion by transferring \(SOC\) from and to natural vegetation (Sect. \ref{sec:carbontransfer}),
(3) we estimate \(SOC\) stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \ref{sec:totalsoc}).
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \ref{sec:initsoc}).

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the steady-state method for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model (\citep{parton_analysis_1987}). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below- and above-ground residues left or returned to the field (see Sect. \ref{sec:residues}) and manure inputs (see Sect. \ref{sec:livstmanure}); for natural vegetation litterfall (including fine root turnover) (\citep{schaphoff_lpjml4_2018}) is the only source of carbon inflow to the soil. Following the IPCC guidelines \citep{calvo_buendia_ipcc_2019}, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective \(SOC\) sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \ref{tab:datasourceinputs}.

  \begin{table*}[h]
  \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
  \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.25\textwidth} p{0.40\textwidth}}
  \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
  \middlehline
  \multirow{3}{*}{Cropland} & above-ground residues & \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                      \cite{faostat_faostat_2016}, \\
                                                                      \cite{schaphoff_lpjml4_2018} (LPJmL),\\
                                                                      \cite{weindl_livestock_2017} \end{minipage}} & 
                                                      \multirow{3}{*}{\begin{minipage}[t]{1\columnwidth}\raggedright
                                                                      LG:C generic values according to Table 5.5B\\
                                                                      in \cite{calvo_buendia_ipcc_2019},\\
                                                                      crop-type specific N:C-ratios from
                                                                      \cite{bodirsky_mrcommons_2020}\end{minipage}} \\
                            & below-ground residues &  &  \\
                            & manure                &  &  \\ 
                            \hline
  Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018} (LPJmL) 
                            & \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut 
                                \cite{calvo_buendia_ipcc_2019} (CENTURY parameterization) \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

@@

Note that the absolute changes \(\Delta SOC\) can be also interpreted as the SOC debt (\citep{sanderman_soil_2017}) due to human cropping activities; whereas relative changes \(F^{\mathrm{SCF}}\) can be considered stock change factors as defined within the IPCC guidelines of 2006 \citep{eggleston_ipcc_2006}. Moreover, \(\Delta SOC\) is equivalent to the negated cumulative SOC component of human land-use change emissions (\citep{pugh_simulated_2015}).

@@

We compile country-specific FAO production and cropland statistics \citep{faostat_faostat_2016} to a harmonized and consistent data set. The data is prepared in 5-year time steps from 1965 to 2010, which restricts our analysis to the time span from 1975 to 2010 (after a spin-up phase from 1901-1974). For all the following data, if not declared differently, we interpolate values linearly between the time steps and keep them constant before 1965.

@@

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL \citep{schaphoff_lpjml4_2018} as well as half-degree cropland patterns (see Sect. \ref{sec:landuse}). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production \citep{faostat_faostat_2016}. By using physical cropland areas in combination with harvested areas, we account for multiple cropping systems as well as for fallow land.

@@

Following the IPCC guidelines, we split the harvest index into a production (\(HI^{\mathrm{prod}}\)) and an area dependent (\(HI^{\mathrm{area}}\)) fraction \citep{eggleston_ipcc_2006}. Deviating from \citep{bodirsky_n2o_2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all \(BGR\) are left in the soil, whereas \(AGR\) can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

A country-specific fixed share of the \(AGR\) is assumed to be burned on field depending on the per-capita income of the country. Following \citep{smil_nitrogen_1999}) we assume a burn share of \(25\%\) for low-income countries according to World Bank definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80--90\% of the carbon in the crop residues burned in the fields is lost within the combustion process \citep{eggleston_ipcc_2006}.

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in \citep{bodirsky_mrcommons_2020}). Residue feed demand for five different livestock groups is based on country-specific feed baskets (see \citep{weindl_livestock_2017}), that differentiate between the residue groups and take available \(AGR\) biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5\% and a fuel-share of 10\% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining \(AGR\) as well as all \(BGR\) are expected to be left on the field. We limit high residue return rates to at most \(10\unit{tC\,ha}^{-1}\) in order to correct for outliers.

@@

To disaggregate country level FAOSTAT livestock production data to half-degree resolution, we use the following rule-based assumptions, drawing from the approach of \citep{robinson_mapping_2014} and applying feed basket assumptions based on a revised methodology from \citep{weindl_livestock_2017}. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
Due to great feed demand of ruminants, we assume that ruminant livestock is located where the production of feed occurs to minimize transport of feed. We distinguish between grazed pasture, which is converted into livestock products in extensive systems, and primary-crop feed stuff, which we consider to be consumed in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to distinguish between intensive and extensive production systems. For low-income countries, we assume only extensive production systems. We locate them according to the share of built-up areas based on the assumption that these animals are held in subsistence or small-holder farming systems with a high labor-per-animal ratio. Intensive production associated with high-income countries, is distributed within a country using the share of primary-crop production, assuming that feed availability is the most determining factor for livestock location. For middle-income countries we split the livestock production into extensive and intensive systems based on the per-capita income.

Manure production and management is based on a revised methodology of \citep{bodirsky_n2o_2012} and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets \citep{weindl_livestock_2017}, assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see \citep{calvo_buendia_ipcc_2019}).
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25\% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing.
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates depending on the different systems and the associated nitrogen loss rates as specified in \citep{bodirsky_n2o_2012} (see Sect. \ref{append:TableclossAWMS} for more details). We limit high application shares at \(10\unit{tC\,ha}^{-1}\) to correct for outliers.

@@

The LUH2v2 \citep{hurtt_harmonization_2020} data set provides irrigated fractions for their cropland subcategories. We sum up irrigation area shares for all crop groups within a grid cell, and calculate the water effect coefficient \(wat\) on decay rates using these shares to compute the weighted mean between rainfed and irrigated \(wat\) factors. As a result \(wat\) is the same for all crop groups within a grid cell. Furthermore, we suppose the irrigation effect to be present for all 12 months of a year in a grid cell including irrigated areas, since we do not have consistent crop group specific growing periods available. This will lead to an overestimation of the irrigation effect. We expect, however, water limitations to be a minor problem during the off-season in temperature limited cropping regions, causing our assumption to not dramatically overestimate the moisture effects. In tropical, water-limited cropping areas, irrigated growing periods might even span over the whole year.

@@

To single out the impact of tillage practices, residue and manure inputs, we defined scenario with constant values for these three drivers: In the \textit{constTillage} scenario the adoption of no-tillage practices are neglected (adoption starts in 1974 according to the available data set). The \textit{constResidues} and the \textit{constManure} scenario assume constant input rates from residues resp. manure (in \(\tfrac{t}{ha}\)) at the level of 1975 onwards. Within the \textit{constResidue} scenario at different effects overlay each other: yields and with them residue biomass increase due to productivity gains; rates of residue left or returned to fields are raising; and shifts of cropping pattern change the amount of residue biomass due to crop-group specific harvest index values. The \textit{constManagement} combines all three scenarios \textit{constTillage}, \textit{constResidues} and \textit{constManure}.

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short LG:C) and nitrogen-to-carbon (short N:C) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines (\citep{calvo_buendia_ipcc_2019}), but for woody biomass we used the CENTURY configuration file (\citep{century_model_2000}) for trees as parameterization were not given. Whereas LG:C ratios ranges between \(0.09-0.35\) with an average of \(0.24\) (default value) for different tree compartments and different tree types, N:C ratios range between \(0.001-0.05\) with an average of \(0.015\) (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table \ref{tab:scenlitterpnv}.

 \begin{table*}[h]
 \caption{Scenarios for sensitivity analysis on litter parameterization}
 \begin{tabular}{l l l l}
 \tophline
  Scenario name & plant type & N:C & LG:C \\
  \middlehline
  \textit{LitterPNV-Mean} & mix over all tree types & 0.015  & 0.25 \\
  \textit{LitterPNV-Grass} & perennial grasses     & 0.029  & 0.11 \\
  \textit{LitterPNV-AASS}  & Arctic alpine subshrub           & 0.022  & 0.22 \\
  \textit{LitterPNV-BLEAF} & Temperate broadleaf evergreen    & 0.013  & 0.27 \\
  \textit{LitterPNV-BRCF}  & Boreal coniferous forest         & 0.013  & 0.17 \\
  \textit{LitterPNV-MTCF}  & Maritime coniferous forest       & 0.0094 & 0.22 \\
  \textit{LitterPNV-TREF}  & Tropical evergreen forest        & 0.016  & 0.30 \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:scenlitterpnv}
 \end{table*}

@@

This study shows that spatially explicit and time-variant historical agricultural management data changes estimations of states and trends of global SOC and SOC debt considerably compared to the often used constant management assumptions. This result remains robust to variations of central model parameters and variations in the initialization of SOC stocks. While land cover change has depleted SOC stocks and increased the SOC debt, our analysis points out that the high increases in agricultural productivity may have even led to a net reduction of the SOC debt since 1975.
